Neural Connectome of the Ctenophore Statocyst

### 2025-05-16 manuscript/Manuscript.qmd
MS, refs: update

---
title: "Neural Connectome of the Ctenophore Statocyst"
subtitle: |
  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
  <br><br>  
  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
  <br>
  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
  <br>  
  $^{3}$ Grass Laboratory, Marine Biological Laboratory, Woods Hole, MA 02543, USA  
  <br> 
  $^{4}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
  <br>  
  $^{5}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
  <br> 
  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
  <br>
  $^{7}$ Michael Sars Centre, University of Bergen, Norway
  <br><br>  
  **$^{*}$ Correspondence:**  
  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than effector function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior. Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @Tamm_2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

@@

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants.

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).

@@

ANN_Q1-4 formed synapses on balancer cells in all four quadrants (on 7 cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while ANN_Q1Q2 and ANN_Q3Q4 synapsed to balancer cells in their respective quadrants (ANN_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN_Q3Q4 on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4. While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014; @Tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.

@@

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014]. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024]. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015; @roberts2022].

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014b]. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

@@

As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here similarly orchestrates output through dynamic, self-regulating coordination. These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.

>>>>

---
title: "Neural Connectome of the Ctenophore Statocyst"
subtitle: |
  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,7}$, **Lara Kewalow**$^{7}$, **Pawel Burkhardt**$^{6}$, **Gáspár Jékely**$^{1,2,7,*}$  
  <br><br>  
  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
  <br>
  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
  <br>  
  $^{3}$ Grass Laboratory, Marine Biological Laboratory, Woods Hole, MA 02543, USA  
  <br> 
  $^{4}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
  <br>  
  $^{5}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
  <br> 
  $^{6}$ Michael Sars Centre, University of Bergen, Norway
  <br>
  $^{7}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
  <br><br>  
  **$^{*}$ Correspondence** <br>   
  ![](pictures/emails.png){height=20}<br>  
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than sensory-motor function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise1973a; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior. Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2015]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

@@

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants.

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral organ. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973b]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).

@@

ANN_Q1-4 formed synapses on balancer cells in all four quadrants (on 7 cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while ANN_Q1Q2 and ANN_Q3Q4 synapsed to balancer cells in their respective quadrants (ANN_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN_Q3Q4 on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4. While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@tamm2014b; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.

@@

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013; @moroz2015]. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024]. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015;@roberts2022].

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014c]. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances \[[@waggett1999; @colin2010; @courtney2020]\]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

@@

As Jager et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here similarly orchestrates output through dynamic, self-regulating coordination. These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.
